---
title: "EAE Analysis for Continuous QTVs"
author: Anna L Tyler, modified by Lauren A. Downs
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  html_document:
    code_folding: hide
    collapsed: no
    toc: yes
    toc_float: yes
bibliography: eae.bib
---

## Introduction
The purpose of this workflow is to build on the analysis done by Dr. Emily Nelson, 
Anna Tyler, and Dr. Dimitry Krementsov on a panel of Collaboraitve Cross mice that 
were used to model the heterogeneity of Multiple Sclerosis (MS) using [experimental autoimmune 
enchephalomyelitis (EAE)](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3229753/), a well established mouse model of the disease.

In this model, mice are injected with CNS antigens which cause
development of autoimmunity which results in clinical symptoms (weight loss, ataxia, paralysis) as well as measurable pathology in the brain and spinal cord.
Previously, the Krementsov lab has identified multiple types of EAE that mirror
the human experience better than the standard model, including potential models for different subtypes of MS. Similarly, potentially novel patterns were seen in the histopathology of some strains used in this experiment.

In this study, we will try to map genetic variants that 
are associated with weight change from EAE. Previous data describing the severity of clinical symptoms, known as a Cummulative Disease Score (CDS), will be used as control. Weight change is used because this is a continuous variable and thus will likely map differently than a more binned or discrete metric.

For reference, [Emily's paper](https://pmc.ncbi.nlm.nih.gov/articles/PMC11601571/) provides more background and the foundation for this project. 

### Collecting and Processing Data

Using the files that Anna had previously compiled as well as some of Emily Nelson's data, I made a new 'September 2025' CDS file where I added a column for average weight change and average weight loss (average weight change * -1). These are percentages calculated based on the weight of a mouse over the course of the experiment 


Note: Any code that was edited or made with ChatGPT will be noted within in-line comments.

```{r get_arguments}
rm(list = ls())

#===============================================================
#These are the hard-coded parameters used for the CDS Analysis
#===============================================================
#args <- commandArgs(trailingOnly=T)
#b6_correction <- as.logical(args[1]) #Use B6 score to do batch correction
#batch_as_covar <- as.logical(args[2]) #use batch as a covariate
#rank_norm <- as.logical(args[3]) #whether to rank Z normalize adjusted scores.
#===============================================================



#===============================================================
#hard code the final decisions on corrections and normalizations
#set final_param to FALSE to use the arguments again
#===============================================================
final_param = TRUE

if(final_param){
    b6_correction <- FALSE
    batch_as_covar <- TRUE
    rank_norm <- TRUE
}
#===============================================================

genotypes <- c("MRCA", "sequenced")
genotypes <- genotypes[2]

if(is.na(b6_correction) || length(b6_correction) == 0){
b6_correction = FALSE; batch_as_covar = FALSE; rank_norm = FALSE
}

if(b6_correction && batch_as_covar){stop("Set either b6_correction or batch_as_covar to TRUE, not both.")}

alpha <- c(0.15, 0.2)
```

```{r load_code}
library(here)

all.fun <- list.files(here("Code"), full.names = TRUE, pattern = ".R")
for(i in 1:length(all.fun)){source(all.fun[i])}

overwrite_previous <- FALSE # make this true whenever you add new data that you want fully analyzed; this WILL take longer and may not work on a low powered machine
add_founder_genotypes <- FALSE

#strains identified by remove_strains are excluded from all 
#manipulations including normalization and batch corrections.
#justficiations are as follows:
#CC027 - all but 2 were homozygous for the wrong genotype at the H2 locus
#NOD and 129 - are not used for mapping, and were not included in all
#batches, so we cannot use them for normalization
#remove_strains <- NULL
remove_strains <- c("CC027", "NOD", "129") 
#remove_strains <- c("CC051", "CC059") #closely related
#remove_strains <- c("CC051") #pick one of them to keep

cc.color.scheme = "original"
#cc.color.scheme = "new"

#for CDS
exclude.zeros = FALSE #if true, we exclude animals with a CDS of 0
disease.thresh = 0 #maximum disease score considered disease-free

cds.file.date = "March_2026" #date for CDS csv file to use in format month_year


nperm = 1000

if(nperm > 0){
    run_permutations <- TRUE
}else{
    run_permutations <- FALSE
}
```

## Parameters

```{r param_text}
if(add_founder_genotypes){
    founders.included = "yes"
}else{
    founders.included = "no"
}

if(length(remove_strains) > 0){
    strains.removed <- paste(remove_strains, collapse = ", ")
}else{
    strains.removed = "none"
}

if(!exclude.zeros){
    zero.text <- "Zeros are included in CDS"
    zero.treatment <- "included with"
}else{
    zero.text <- paste("Animals with scores of", disease.thresh, "and below are excluded from CDS.")
    zero.treatment <- "excluded from"
}

if(b6_correction){
    correction.text <- "Batch corrections were made using B6 CDS."
    correction.label <- "B6-batch"
}
if(batch_as_covar){
    correction.text <- "Batch corrections were done using batch as a covariate."
    correction.label <- "covar-batch"
}
if(!b6_correction && !batch_as_covar){
    correction.text <- "No batch corrections were done."
    correction.label <- "no-batch"
}

if(rank_norm){
    norm.text <- "Rank Z normalization was performed."
    norm.label <- "rank-norm"
}else{
    norm.text <- "Rank Z normalization was not performed."
    norm.label <- "no-norm"
}

file.label <- paste(correction.label, norm.label, genotypes, sep = "_")
```

The parameters for this run are as follows (based on previous CDS analysis):

* Genotypes used: `r genotypes`

* Founders included: `r founders.included`

* CC Strains removed: `r strains.removed`

* Treatment of zeros: `r zero.text`

* Batch correction: `r correction.text`

* Normalization: `r norm.text`

```{r results_dir}
results.dir <- here("Results", paste("Zeros", strsplit(zero.treatment, " ")[[1]][1], file.label, sep = "_"))
if(!file.exists(results.dir)){dir.create(results.dir)}
```

```{r load_libraries, warning = FALSE, message = FALSE, error = FALSE}
all.packages <- c("pheatmap", "qtl2", "ggplot2", "abind", "igraph", "knitr", "parallel",
    "stringr", "grid")
load_libraries(all.packages)
cores = NULL #detectCores()-1 is the alternative code if the machine you are using has cores
```

Read in phenotype data.

```{r load_data}
if(cc.color.scheme == "original"){
    strain.colors <- CCorigcolors
}else{
    strain.colors <- CCcolors
}

cds <- read.csv(here("Data", paste0("CDS_", cds.file.date, ".csv")), strip.white = TRUE)

#take out any slashes in strain names
cds[,"Strain"]  <- gsub("/", "", cds[,"Strain"], fixed = TRUE)


```

This run uses the CDS file from `r gsub("_", " ", cds.file.date)`.


### Remove excluded strains

If any strains need to be removed from the analysis, do 
that now before any numbers are manipulated.


```{r remove_strains}
#remove any empty rows that were inserted to visually separate
#the founders from the rest of the strains
cds.data.idx <- which(!is.na(cds[,"ID"]))
cds <- cds[cds.data.idx,]


#remove specified strains from cds
if(length(remove_strains) != 0){
    remove.cds <- unlist(sapply(remove_strains, function(x) which(cds[,"Strain"] == x)))
    cds <- cds[-remove.cds,]
}


```


```{r batches}
## CDS
num.cds <- as.matrix(sapply(cds[,c("Global.CDS", "Avg.Weight.Change", "Avg.Weight.Loss", "Weight.D0")], as.numeric)) #had to change this with Chat to make the values numeric
rownames(num.cds) <- cds[,"ID"]
#set all scores equal to or lower than the disease threshold
#to zero
if(exclude.zeros){
    num.cds[which(num.cds <= disease.thresh)] <- 0
}

factor.cds <- as.matrix(cds[,c("Strain", "Sex")])
u_batch <- unique(cds[,"EXP"])

#remove animals in batch 170 - these were validation, not part of the original experiment
to.remove = "X170" 
keep.batch <- which(u_batch != "X170")
u_batch <- u_batch[keep.batch]

keep.idx <- which(cds[,"EXP"] != "X170")
cds <- cds[keep.idx,]
num.cds <- num.cds[keep.idx,]
factor.cds <- factor.cds[keep.idx,]

batch.mat <- matrix(0, nrow = nrow(cds), ncol = length(u_batch))
colnames(batch.mat) <- u_batch
for(b in 1:ncol(batch.mat)){
    batch.mat[which(cds[,"EXP"] == u_batch[b]),b] <- 1
}
cds.covar <- as.matrix(cbind(batch.mat, cds[,"Age.at.D0",drop=FALSE]))

pheno.strains <- cds[,"Strain"]
eae.strains <- unique(pheno.strains)


```

## Samples by strain

The following heatmaps show how many individuals are in 
each strain by sex combination per CDS.

### CDS

```{r sample_counts_cds, fig.width = 8, fig.height = 1.5}
count.table <- table(cds[,c("Sex", "Strain")])
strain.order <- c(3:ncol(count.table), 1,2)
par(mar = c(1,4,4,2))
imageWithText(count.table[,strain.order], use.pheatmap.colors = TRUE, 
    col.text.shift = -1.4, row.text.shift = -2, col.text.adj = 0,
    col.text.cex = 0.7, cex = 0.8, row.text.adj = 0.5)
```


## Data Distributions 

### CDS Distribution {.tabset .tabset-fade .tabset-pills}

Each day of the study, each mouse is given a disease CDS 
score between 0 and 5, with lower numbers indicating less 
disease. The cumulative disease score (CDS) for a given mouse
is the sum of these scores over the course of the study.

Similarly, the weight of each mouse was measured since weight loss can be a sign of increased disease severity in mouse models. These measurements were then averaged for each mouse and a resulting average weight Loss metric was created.

The distributions of three scores and the average weight Loss are shown below. Classical
is the classical presentation of paralysis used in the field.
AR (axial rotary) scores an alternate form of paralysis.
Global is the sum of the Classical and AR scores and gives
an overall score of disease regardless of type. 
Average weight Loss is, as the title implies, a metric describing whether the mice lost or gained weight overall

```{r CDS zeros, fig.width = 8, fig.height = 8}
par(mfrow = c(2,2))
for(i in 1:ncol(num.cds)){
    hist(num.cds[,i], breaks = 50, main = colnames(num.cds)[i], 
    xlab = colnames(num.cds)[i])
}
```

There are many mice that never display measurable symptoms
and have cumulative disease scores of 0. This kind of floor
effect causes problems with statistical tests that assume
a normal distribution.

If exclude.zeros is set to TRUE, these 0 values will be
excluded from CDS mapping.

In this analysis `r zero.text`.

The plots below show the CDS distribution by strain.

```{r cds, results = "asis", fig.width = 10, fig.height = 5}

plot_qtl_by_strain <- function(scores, strain.labels, exclude.zeros = TRUE,
    ymin = NULL, ymax = NULL, plot.label = ""){
    
    u_strain <- unique(strain.labels)
    strain.idx <- lapply(u_strain, function(x) which(strain.labels == x))

    if(exclude.zeros){
        nonzero.idx <- which(scores != 0)
    }else{
        nonzero.idx <- 1:length(scores)
    }
    strain.cds <- lapply(strain.idx, function(x) scores[intersect(nonzero.idx, x)])
    names(strain.cds) <- u_strain
    strain.order <- order(sapply(strain.cds, mean))

    if(is.null(ymax)){ymax = max(scores, na.rm = TRUE)}
    if(is.null(ymin)){
        ymin = min(scores, na.rm = TRUE)
        #if(ymin > 0){ymin <- 0}
    }

    #quartz(width = 10, height = 5)
    layout(matrix(c(1,2), ncol = 2), width = c(1, 0.5))
    par(mar = c(4,4,4,0))
    stripchart(strain.cds[strain.order], vertical = TRUE, pch = 16, 
        col = "gray", method = "jitter", las = 2, ylim = c(ymin, ymax),
        main = paste(plot.label), ylab = paste(plot.label, "values"))
    a <- hist(unlist(strain.cds), plot = FALSE)
    par(mar = c(4,0,4,2))
    plot.new()
    plot.window(xlim = c(0, max(a$counts)), ylim = c(ymin, ymax))
    segments(x0 = rep(0, length(a$counts)), x1 = a$counts, y0 = a$mids, 
        lwd = 10, col = "gray")
    #text(x = a$counts+2, y = a$mids, labels = a$counts, adj = 0, cex = 0.7)
    mtext("Individual Counts", side = 3)
    axis(1)

    invisible(strain.cds)
}

for(cd in 1:ncol(num.cds)){
    cat("####", colnames(num.cds)[cd], "\n")
    temp.vals <- plot_qtl_by_strain(scores = num.cds[,cd], 
        strain.labels = factor.cds[,"Strain"], exclude.zeros = exclude.zeros,
        ymax = max(num.cds[,1:3]), plot.label = colnames(num.cds)[cd]) # didn't like max() for all of them so using Chat for syntax reset it to max the former 3 columns
    cat("\n\n")
}
```


## Batch Effects

The individuals were run in multiple batches, which can 
cause systematic biases in the data. We need to correct
any batch effects that we find before we map.

The following heatmaps show how the different 
strains and sexes were represented in different 
batches. The males and females are shown separately.

### CDS

```{r CDS batch_assortment, fig.width = 5, fig.height = 7}
batch.list <- vector(mode = "list", length = length(u_batch))
for(b in 1:length(u_batch)){
    batch.idx <- which(cds.covar[,b] == 1)
    batch.mat <- matrix(0, nrow = length(eae.strains), ncol = 2)
    rownames(batch.mat) <- eae.strains
    batch.table <- table(data.frame(factor.cds[batch.idx,]))
    batch.mat[match(rownames(batch.table), eae.strains),] <- batch.table
    colnames(batch.mat) <- paste(colnames(cds.covar)[b], colnames(batch.table), sep = "_")
    batch.list[[b]] <- batch.mat
}
all.batch.counts <- Reduce("cbind", batch.list)
imageWithText(all.batch.counts, use.pheatmap.colors = TRUE, col.text.shift = -38.5,
col.text.rotation = 45, col.text.adj = 0, cex = 0.7)
```

The following boxplots show the batch values for the B6 
animals. These show that the batches had an increase in 
CDS across the batches.

```{r CDS_b6_batch, fig.width = 8, fig.height = 8, eval = FALSE}
b6.idx <- which(factor.cds[,1] == "C57BL6")

if(exclude.zeros){
    nonzero.idx <- lapply(1:ncol(num.cds), function(x) which(num.cds[,x] != 0))
}else{
    nonzero.idx <- lapply(1:ncol(num.cds), function(x) 1:nrow(num.cds))
}

b6.batch.idx <- apply(cds.covar[,1:(ncol(cds.covar)-1)], 2, 
    function(x) intersect(which(x == 1), b6.idx))

par(mfrow = c(2,2))
b6.by.batch <- vector(mode = "list", length = ncol(num.cds))
names(b6.by.batch) <- colnames(num.cds)
for(cd in 1:ncol(num.cds)){
    #get batch values for this CDS
    vals.by.batch <- lapply(1:length(b6.batch.idx), 
        function(x) num.cds[intersect(b6.batch.idx[[x]], nonzero.idx[[cd]]),cd])
    names(vals.by.batch) <- u_batch
    b6.by.batch[[cd]] <- vals.by.batch
    boxplot(b6.by.batch[[cd]], main = colnames(num.cds)[cd])    
}

```


### Batch correction procedure

The B6 batch correction procedure is as follows:

1. Calculate the overall mean CDS for B6 mice 
    across the entire experiment
2. Calculate the mean CDS B6 mice in each batch.
3. Adjust all mice in each batch based on the
    difference in means in the B6. Subtract off 
    the mean deviation, and divide by the standard
    deviation.
4. Male and female B6 mice do not have a sex 
    difference in CDS, and adjusting by the 
    two sexes separately could inadvertently 
    introduce a sex effect into the other strains,
    so we use overall means rather than sex-specific 
    means.

The batch correction procedure using batch as a covariate is as follows:

1. Create a binary matrix indicating whether each individual
    was run in each batch (1) or not (0).
2. Regress this matrix out of the CDS matrix.


#### Remove batch effects from CDS{.tabset .tabset-fade .tabset-pills}

```{r adj_cds}
#choose which phenotypes to batch 
#correct using their indices.
cds.idx <- c(1,2,3,4) 
```

Adjust for batch. We adjusted the following traits for batch effects:
`r cat(colnames(num.cds)[cds.idx], sep = "\n")`

```{r no_adj_cds}
adj.cds <- num.cds 
```

```{r adj_batch_covar_cds}
if(batch_as_covar){
    adj.cds <- adjust(num.cds, cds.covar[,1:ncol(cds.covar)-1])
}
```

```{r mean_sd_cds}

if(b6_correction){
    batch.mean <- sapply(b6.by.batch, function(x) sapply(x, mean))
    batch.sd <- sapply(b6.by.batch, function(x) sapply(x, sd))
    batch.sd[which(is.na(batch.sd))] <- 1

    #for each batch
    for(b in c(1:(ncol(cds.covar)-1))){
        batch.idx <- which(cds.covar[,b] == 1)
        for(j in cds.idx){ #only correct classical and global
            if(exclude.zeros){
                    nonzero.idx <- which(num.cds[,j] != 0)
                }else{
                    nonzero.idx <- 1:nrow(num.cds)
                }
            adj.idx <- intersect(batch.idx, nonzero.idx)
            adj.cds[adj.idx,j] <- (num.cds[adj.idx,j] - batch.mean[b,j])/batch.sd[b,j]
        }
    }
    
}
```




### Age at D0 {.tabset .tabset-fade .tabset-pills}

Another one of our covariates is Age.at.D0.

The following boxplots show that age varies significantly 
by batch.

#### CDS

```{r age_by_batch_cds}
age <- cds.covar[,ncol(cds.covar)]
age.by.batch <- lapply(1:(ncol(cds.covar)-1), function(x) age[which(cds.covar[,x] == 1)])
p.val <- aov.by.list(age.by.batch)$"Pr(>F)"[1]

boxplot(age.by.batch, main = paste("Age at D0 by batch\n", "p =", signif(p.val, 2)))
stripchart(age.by.batch, vertical = TRUE, method = "jitter", pch = 16, 
    col = "#756bb1", add = TRUE)
```

The following plots show the relationship between Age.at.D0
and each CDS after adjusting for batch.

```{r age_at_day0_cds, results = "asis", fig.width = 8, fig.height = 8}

u_age <- unique(cds.covar[,"Age.at.D0"])
age.idx <- lapply(u_age, function(x) which(cds.covar[,"Age.at.D0"] == x))
cds.by.age <- lapply(age.idx, function(x) adj.cds[x,,drop=FALSE])

par(mfrow = c(2,2))
for(i in 1:4){
    just.one <- lapply(cds.by.age, function(x) x[,i])
    x.vals <- lapply(1:length(just.one), function(x) rep(u_age[x], length(just.one[[x]])))
    x <- unlist(x.vals)
    y <- unlist(just.one)
    if(exclude.zeros){
        nonzero.idx <- which(y != 0)
    }else{
        nonzero.idx <- 1:length(y)
    }

    plot.with.model(x[nonzero.idx], y[nonzero.idx], xlab = "Age at D0", 
        ylab = colnames(num.cds)[i], main = colnames(num.cds)[i])
}
```


```{r adjust_age_cds, eval = FALSE}
#adjust for age
for(i in 1:ncol(adj.cds)){
    if(exclude.zeros){
        nonzero.idx <- which(num.cds[,i] != 0)
    }else{
        nonzero.idx <- 1:nrow(num.cds)
    }
    adj.cds[nonzero.idx,i] <- adjust(adj.cds[nonzero.idx,i,drop=FALSE], 
        cds.covar[nonzero.idx,"Age.at.D0",drop=FALSE])
}
```

```{r write_adj_cds}
#write out the results of whatever adjustments we have done
write.table(adj.cds, file.path(results.dir, "Adj_CDS.csv"), sep = ",", quote = FALSE)
```

Rank Z normalize the CDS scores if this is specified.

```{r no_norm_cds}
norm.cds <- adj.cds
colnames(norm.cds) <- gsub(".CDS", "", colnames(norm.cds)) #take off CDS in labels, to save room in plots
```

```{r norm_cds}
if(rank_norm){
    for(i in 1:ncol(norm.cds)){
        if(exclude.zeros){
            nonzero.idx <- which(num.cds[,i] != 0)
        }else{
            nonzero.idx <- 1:nrow(num.cds)
        }
        norm.cds[nonzero.idx,i] <- rankZ(adj.cds[nonzero.idx,i])
    }
}
colnames(norm.cds) <- gsub(".CDS", "", colnames(norm.cds)) #take off CDS, to save room in plots

#write out normzliaed values
write.table(norm.cds, file.path(results.dir, "Norm_CDS.csv"), sep = ",", quote = FALSE)

```

## CC genotypes

I downloaded the CC genotypes from the UNC website:
http://csbio.unc.edu/CCstatus/CCGenomes/#genotypes

I downloaded the Most Recent Common Ancestor file. 
This has marker names from the mega muga. I downloaded the
mega muga marker map from the Jackson Laboratory 
[resource for CC/DO mice](https://www.jax.org/research-and-faculty/genetic-diversity-initiative/tools-data/diversity-outbred-reference-data)


```{r MRCA_genotypes, warning = FALSE}
cc.geno <- readRDS(here("Data", "MRCA10_modelprobs_trim.rds"))

if(genotypes == "MRCA"){

    marker.info.file <- here("Data", "mouse", "mm_uwisc_v1.csv")
    marker.info <- read.csv(marker.info.file)

    #convert the map and the genoprobs into a qtl2 object
    cross_obj <- array2qtl2(cc.geno, marker.info, chr = c(1:19, "X"), chr.col = "chr", 
        pos.col = "bp_mm10", crosstype = "risib8")

    cc.genoprobs <- cross_obj$genoprobs
    map <- cross_obj$map

    K <- calc_kinship(cc.genoprobs, "loco")
}
```


```{r sequenced_genotypes}

if(genotypes == "sequenced"){
    map.file <- here("Data", "map.RDS")
    new.geno.file <- here("Data", "genoprobs.RDS")
    k.file <- here("Data", "kinship_loco.RDS")

    if(!file.exists(map.file)){
        file <- paste0("https://raw.githubusercontent.com/rqtl/", "qtl2data/master/CC/cc.zip")
        cc <- read_cross2(file)
        map <- cc$pmap
        saveRDS(map, map.file)
        
        cc.official <- rownames(cc$geno[[1]])
        cc.alias <- sapply(strsplit(cc.official, "/"), function(x) x[1])
        write.table(cbind(cc.official, cc.alias), here("Data", "CC_names.txt"), sep = "\t", 
            quote = FALSE, row.names = FALSE, col.names = FALSE)
        
        cc.genoprobs <- calc_genoprob(cc, cores = cores)    
        new.genoprobs <- cc.genoprobs

        #add strains 83 and 84 to the genoprobs from the MRCA genotypes
        new.strains <- c("CC083", "CC084")
        new.strain.idx <- match(new.strains, rownames(cc.geno))
        for(i in 1:length(cc.genoprobs)){
            chr.markers <- dimnames(cc.genoprobs[[i]])[[3]]
            common.markers <- intersect(chr.markers, dimnames(cc.geno)[[3]])
            orig.idx <- match(common.markers, dimnames(cc.genoprobs[[i]])[[3]])
            new.idx <- match(common.markers, dimnames(cc.geno)[[3]])
            new.chr <- abind(cc.genoprobs[[i]][,,orig.idx], cc.geno[new.strain.idx,,new.idx], along = 1)
            new.genoprobs[[i]] <- new.chr
        }
        saveRDS(new.genoprobs, new.geno.file)

        #overwrite the original objecct
        cc.genoprobs <- new.genoprobs

        K <- calc_kinship(new.genoprobs, "loco")
        saveRDS(K, k.file)
    }else{
        map <- readRDS(map.file)
        cc.genoprobs <- readRDS(new.geno.file)
        K <- readRDS(k.file)
    }
}
```

If specified, add genoptypes for 129, NOD, and B6, so we 
can use them in mapping.

```{r add_other_strains}
if(add_founder_genotypes){
    founder_fun <- function(strain.name){
        founder.genotypes <- rep(0, length(strain.colors))
        founder.idx <- which(names(strain.colors) == strain.name)
        founder.genotypes[founder.idx] <- 1
        return(founder.genotypes)
    }

    strain.129 <- founder_fun("129")
    strain.b6 <- founder_fun("B6")
    strain.nod <- founder_fun("NOD")

    mapping.genoprobs  <- cc.genoprobs
    for(i in 1:length(cc.genoprobs)){
        ind.129 <- matrix(strain.129, nrow = ncol(cc.genoprobs[[i]]), 
            ncol = dim(cc.genoprobs[[i]])[3])
        ind.b6 <- matrix(strain.b6, nrow = ncol(cc.genoprobs[[i]]), 
            ncol = dim(cc.genoprobs[[i]])[3])
        ind.nod <- matrix(strain.nod, nrow = ncol(cc.genoprobs[[i]]), 
            ncol = dim(cc.genoprobs[[i]])[3])
        new.chr <- abind(cc.genoprobs[[i]], ind.129, ind.b6, ind.nod, along = 1)
        rownames(new.chr)[tail(1:length(rownames(new.chr)), 3)] <- c("129", "C57BL6", "NOD")
        mapping.genoprobs[[i]] <- new.chr
    }

    #recalculate K for the new genoprobs.
    mapping.K <- calc_kinship(mapping.genoprobs, "loco")
    mapping.strains <- eae.strains
}else{
    mapping.genoprobs <- cc.genoprobs
    mapping.K <- K
    mapping.strains <- eae.strains[grep("CC", eae.strains)]
}
```

## Allele Frequencies {.tabset .tabset-fade .tabset-pills}

The following plots show the frequencies of each founder 
allele across each chromosome. Regions in which the allele
frequencies are imbalanced may not be good for mapping. Similar to genotypes this should be the same for CDS.

```{r allele_freq, results = "asis", fig.width = 8, fig.height = 4}
allele_freq <- lapply(mapping.genoprobs, function(x) apply(x, 3, colMeans))
for(ch in 1:length(allele_freq)){
    cat("### Chr", names(map)[[ch]], "\n")
    layout(matrix(c(1,2), nrow = 1), widths = c(1,0.3))
    par(mar = c(4,4,4,0))
    plot.new()
    plot.window(xlim = c(0, max(map[[ch]])), ylim = c(0,0.4))
    for(a in 1:nrow(allele_freq[[ch]])){
        points(map[[ch]][dimnames(mapping.genoprobs[[ch]])[[3]]], 
        allele_freq[[ch]][a,], col = strain.colors[a], type = "l", lwd = 2)
    }
    abline(h = 0)
    axis(1)
    axis(2)
    mtext(paste("Chr", names(map)[[ch]]))
    plot.new()
    plot.window(xlim = c(0,1), ylim = c(0,1))
    par(mar = c(4,0,4,0))
    legend(x = 0, y = 1, lty = 1, col = strain.colors, legend = names(strain.colors), 
        horiz = FALSE, bg = "white", lwd = 2, adj = 0)
    cat("\n\n")
}
```


## Original Processing

### Normalized Scores {.tabset .tabset-fade .tabset-pills}

The following plots show the distribution of each 
disease score after adjustments and normalization.

```{r norm_dist_CDS, fig.width = 10, fig.height = 5, results = "asis"}
#ymin = min(norm.cds);ymax = max(norm.cds) #all on same axes
ymin = NULL; ymax = NULL #all on independend axes
for(cd in 1:ncol(norm.cds)){
    cat("####", colnames(norm.cds)[cd], "\n")
    plot_qtl_by_strain(scores = norm.cds[,cd], 
        strain.labels = factor.cds[,"Strain"], exclude.zeros = exclude.zeros,
        ymin = ymin, ymax = ymax, plot.label = colnames(num.cds)[cd])
    cat("\n\n")
}

```





### Overall Sex Effect for CDS

```{r sex_diff_cds, fig.height = 8, fig.width = 8}
#quartz(width = 8, height = 8)
par(mfrow = c(2,2))
for(i in 1:ncol(norm.cds)){
    if(exclude.zeros){
        nonzero.idx <- which(norm.cds[,i] != 0)
    }else{
        nonzero.idx <- 1:nrow(norm.cds)
    }
    model <- lm(norm.cds[nonzero.idx,i]~as.factor(factor.cds[nonzero.idx,"Sex"]))
    f <- summary(model)$fstatistic
    p <- pf(f[1],f[2],f[3],lower.tail=F)
    boxplot(norm.cds[nonzero.idx,i]~as.factor(factor.cds[nonzero.idx,"Sex"]), 
        ylab = colnames(norm.cds)[i],
        main = paste(colnames(norm.cds)[i], "\np =", signif(p, 2)), xlab = "")
}
```



### Trait Correlations

The following plot shows that global CDS
is dominated by classical disease.
It also shows that CDS in classical 
disease and AR are unrelated to each other. 
Lastly, it shows the correlations between the
average weight Loss and the CDS data which are negative. This makes sense because weight loss is associated with higher disease severity.

```{r trait_cor_cds, fig.width = 8, fig.height = 12}
trait.pairs <- pair.matrix(1:ncol(norm.cds))
par(mfrow = c(3,2))
for(i in 1:nrow(trait.pairs)){
    trait1 <- trait.pairs[i,1]
    trait2 <- trait.pairs[i,2]
    if(exclude.zeros){
        nonzero.idx <- intersect(which(norm.cds[,trait1] != 0), which(norm.cds[,trait2] != 0))
    }else{
        nonzero.idx <- 1:nrow(norm.cds)
    }
    plot.with.model(norm.cds[nonzero.idx,trait1], norm.cds[nonzero.idx,trait2],
        xlab = colnames(norm.cds)[trait1], ylab = colnames(norm.cds)[trait2],
        main = paste(colnames(norm.cds)[trait1], "vs.", colnames(norm.cds)[trait2]))
}
```


### Map CDS with Individuals

We mapped each CDS across the individual mice. 

```{r methods}
if(add_founder_genotypes){
    methods.line <- "Both CC and founder strains were used for mapping."
}else{
    methods.line <- "We used only CC strains for mapping."
}

if(length(remove_strains) != 0){
    remove.line <- paste("We excluded the following strains:", paste(remove_strains, collapse = ", "))
}else{
    remove.line <- "No CC strains were excluded."
}


```

`r methods.line`
`r remove.line`

For the genoprobs and kinship matrices, we copied
the entries for the individuals in each strain and
used indices for row names.

The LOD traces for each trait are shown below along 
with plots for the FDR results. Significance thresholds
from the permutations are also shown on the LOD traces.

```{r expand_matrices_cds}

#remove any strains not used in mapping
mapping.strain.idx <- unlist(sapply(mapping.strains, function(x) grep(x, pheno.strains)))
pheno.strains_cds <- pheno.strains[mapping.strain.idx]

#make a unique ID for each mouse that contains its strain and sex
u_id <- as.vector(apply(cds[mapping.strain.idx,c(1:2,5)], 1, function(x) paste(x, collapse = "_")))

K.idx <- sapply(mapping.strains, function(x) grep(x, rownames(mapping.K[[1]])))
K.num <- sapply(mapping.strains, function(x) length(which(pheno.strains_cds == x)))
expanded.idx <- unlist(lapply(1:length(K.idx), function(x) rep(K.idx[x], K.num[x])))
groups <- names(expanded.idx)

exp.K <- lapply(1:length(mapping.K), function(x) mapping.K[[x]][expanded.idx,expanded.idx])
names(exp.K) <- names(mapping.K)

#create unique row and column names
for(i in 1:length(exp.K)){
    rownames(exp.K[[i]]) <- colnames(exp.K[[i]]) <- 1:length(u_id)
}

expanded.genoprobs <- mapping.genoprobs
for(i in 1:length(expanded.genoprobs)){
    expanded.genoprobs[[i]] <- mapping.genoprobs[[i]][expanded.idx,,]
    rownames(expanded.genoprobs[[i]]) <- 1:length(u_id)
}

#align the phenotypes with the strains
covar <- matrix(as.numeric(as.factor(cds[mapping.strain.idx,"Sex"])), ncol = 1)
rownames(covar) <- 1:length(u_id)

mapping.pheno <- norm.cds[mapping.strain.idx,]
rownames(mapping.pheno) <- 1:length(u_id)

mapping.factors <- factor.cds[mapping.strain.idx,]
mapping.covar <- matrix(as.numeric(as.factor(mapping.factors[,"Sex"])), ncol = 1)
rownames(mapping.covar) <- 1:length(u_id)

```


### Heritability

The heritability of CDS is shown below.

```{r herit_cds, fig.width = 7, fig.height = 5}
overall.k.file <- here("Data", "kinship_overall.RDS")
if(!file.exists(overall.k.file) || overwrite_previous){
    overall.K <- calc_kinship(expanded.genoprobs, "overall")
    saveRDS(overall.K, overall.k.file)
}else{
    overall.K <- readRDS(overall.k.file)
}

ind.herit <- rep(NA, ncol(num.cds))
names(ind.herit) <- colnames(num.cds)
for(i in 1:ncol(num.cds)){
    nonzero.idx <- which(mapping.pheno[,i] != 0)
    ind.herit[i] <- est_herit(mapping.pheno[nonzero.idx,i,drop=FALSE], overall.K)
}

barplot_with_num(round(ind.herit, 2), ylab = "Heritability", xlab = "", 
    ylim = c(0, 0.4), text.shift = 0.09, main = "Trait Heritability Estimates")
```


```{r mapping_fun}
#We have run into problems with extremely small variance
#in genotype probabilities leading to large LOD scores.
#round.digs attempts to solve this by rounding all 
#genotype probabilities to the specified number of 
#significant figures. All CC lines should be either 1 or
#0 at every locus, so 2 digits should be sufficient to 
#get any errant heterozygotes, but also squash out very
#small variations
map_pheno <- function(map.geno, map.pheno, map.K, addcovar = NULL,
    intcovar = NULL, groups = NULL, exclude.zeros = TRUE, 
    round.digs = 2,
    results.file = file.path(results.dir, "scan1_results.RDS")){

    if(!file.exists(results.file) || overwrite_previous){

        #round the genotypes so low-variance markers don't lead
        #to large QTL LOD scores
        rounded.geno <- map.geno
        for(chr in 1:length(rounded.geno)){
            rounded.geno[[chr]] <- round(map.geno[[chr]], round.digs)
        }

        pheno.scans <- vector(mode = "list", length = ncol(map.pheno))
        names(pheno.scans) <- colnames(map.pheno)
        for(ph in 1:ncol(map.pheno)){
            if(exclude.zeros){
                nonzero.idx <- which(map.pheno[,ph] != 0)
            }else{
                nonzero.idx <- 1:nrow(map.pheno)
            }

            pheno.scans[[ph]] <- scan1(rounded.geno, map.pheno[nonzero.idx,ph], 
                addcovar = addcovar, intcovar = intcovar, kinship = map.K)
            #plot(pheno.scans[[ph]], map = map)
        }
        saveRDS(pheno.scans, results.file)
    }else{
        pheno.scans <- readRDS(results.file)
    }
    return(pheno.scans)
}


perm_scans <- function(map.geno, map.pheno, map.k, scan1.results, 
    addcovar = NULL, intcovar = NULL, groups = NULL, nperm = nperm, 
    max.exp.lod = 20, n.founders = 8, exclude.zeros = TRUE, 
    results.file = file.path(results.dir, "perm.RDS"), round.digs = 2){

    #permutations for Matt's FDR method (see https://github.com/j-matthew-mahoney/cdf-fdrci.git)
    #right now this does one phenotype at a time, but we can change that later.
    perm.data <- vector(mode = "list", length = ncol(map.pheno))
    nind = nrow(map.pheno)
    #bin possible LOD scores
    breaks = lod2p(seq(max.exp.lod, 0, -0.01), nind = nind, k = n.founders)

    #round the genotypes to get rid of very small variances that
    #lead to large QTL LOD scores
    rounded.geno <- map.geno
    for(chr in 1:length(rounded.geno)){
        rounded.geno[[chr]] <- round(map.geno[[chr]], round.digs)
    }

    if(!file.exists(results.file)){
        for(ph in 1:ncol(map.pheno)){
            if(exclude.zeros){
                nonzero.idx <- which(map.pheno[,ph] != 0)
            }else{
                nonzero.idx <- 1:nrow(map.pheno)
            }
            if(!is.null(groups)){
                null_stats = scan1_groupperm_stats(rounded.geno, 
                    pheno = map.pheno[nonzero.idx,ph,drop=FALSE], 
                    groups = groups[nonzero.idx],
                    kinship = map.k, nperm = nperm, breaks = breaks, 
                    addcovar = addcovar, intcovar = intcovar,
                    scan1_out = scan1.results[[ph]][,1,drop=FALSE], cores = cores)
            }else{
                null_stats = scan1perm_stats(rounded.geno, 
                    pheno = map.pheno[nonzero.idx,ph,drop=FALSE], 
                    kinship = map.k, nperm = nperm, breaks = breaks,
                    addcovar = addcovar, intcovar = intcovar,
                    scan1_out = scan1.results[[ph]][,1,drop=FALSE], cores = cores)
            }
            perm.data[[ph]] <- null_stats
        }
        saveRDS(perm.data, results.file)
    }else{
        perm.data <- readRDS(results.file)
    }
    return(perm.data)
}

#plots LOD traces with significance thresholds and returns a matrix
#with traits in rows and the different significance thresholds for
#LOD scores in columns corresponding to both standard and FDR-adjusted 
#LOD scores.
plot_fdr_fun <- function(scan.results, perm.results, tab.level = "###", maxy = NULL){

    fdr_summaries <- matrix(NA, nrow = length(scan.results), ncol = 5)
    rownames(fdr_summaries) <- names(scan.results)
    for(ph in 1:length(scan.results)){
        cat(tab.level, names(scan.results)[ph], "\n")
        summary_thresholds = fdr_summarize(null_stats = perm.results[[ph]], 
            fdr_cl = 0.95, geno_sig = alpha, fdr_thresh = c(0.1, 0.2, 0.4), 
            plot = FALSE, quartz = FALSE)

        finite.thresh <- which(is.finite(summary_thresholds))

        if(is.null(maxy)){
            maxy <- max(c(scan.results[[ph]][,1], 
                summary_thresholds[finite.thresh], na.rm = TRUE))*1.1
        }
        par(mar = c(4, 4, 4, 2))
        plot(scan.results[[ph]], map = map, main = names(scan.results)[ph], 
            ylim = c(0, maxy))
        abline(h = summary_thresholds, lty = 1:length(summary_thresholds))
        plot.dim <- par("usr")
        par(xpd = TRUE)
        #text(x = plot.dim[2], y = summary_thresholds[finite.thresh], 
        #    labels = names(summary_thresholds)[finite.thresh], cex = 0.5, adj = 0)
        text(x = plot.dim[2]*1.01, y = summary_thresholds[finite.thresh], 
            labels = alpha, cex = 0.5, adj = 0)

        #text(x = plot.dim[1], y = summary_thresholds[finite.thresh],
        #    labels = summary_thresholds[finite.thresh], cex = 0.7, adj = 1)
        par(xpd = FALSE)
        fdr_summaries[ph,] <- summary_thresholds
        cat("\n\n")
    }
    colnames(fdr_summaries) <- names(summary_thresholds)
    invisible(fdr_summaries)
}

#This function is similar to above, but uses permutation results
#from qtl2's scan1perm, rather than the fdr scan.
plot_pval_fun <- function(scan.results, perm.results, tab.level = "###"){

    sig_summaries <- t(summary(perm.results, alpha = alpha))
    finite.thresh <- which(is.finite(sig_summaries))
    #fdr_summaries <- matrix(NA, nrow = length(scan.results), ncol = length(alpha))
    #rownames(fdr_summaries) <- names(scan.results)
    #colnames(fdr_summaries) <- paste0("FDR ", alpha*100, "%")

    for(ph in 1:length(scan.results)){
        cat(tab.level, names(scan.results)[ph], "\n")

        trait.p <- lod2p(scan.results[[ph]][,1], nind = attr(scan.results[[ph]], "sample_size"), k = 8)
        trait.fdr <- p.adjust(trait.p, "fdr")
        #plot(scan.results[[ph]][,1], trait.fdr, ylim = c(0,1))
        #abline(h = alpha)
        
        fdr.idx <- lapply(alpha, function(x) which(round(trait.fdr, 2) == x))
        fdr.thresh <- unlist(lapply(fdr.idx, function(x) if(length(x) > 0){scan.results[[ph]][x]}else{NA}))
        #fdr_summaries[ph,] <- fdr.thresh

        maxy <- max(c(scan.results[[ph]][,1], 
            sig_summaries[ph,], na.rm = TRUE))*1.1
        par(mar = c(4, 4, 4, 8))
        plot(scan.results[[ph]], map = map, main = names(scan.results)[ph], 
            ylim = c(0, maxy))
        abline(h = sig_summaries[ph,], lty = c(1,2))
        #abline(h = fdr_summaries[ph,])
        plot.dim <- par("usr")
        par(xpd = TRUE)
        text(x = plot.dim[2]*1.005, y = sig_summaries[ph,], 
            labels = colnames(sig_summaries), cex = 0.7, adj = 0)
        par(xpd = FALSE)
        cat("\n\n")
    }
    #sig_summaries <- cbind(sig_summaries, fdr_summaries)
    sig_summaries <- cbind(sig_summaries)
    invisible(sig_summaries)
}

select_threshold <- function(scan.results, fdr_summaries){
    for(ph in 1:length(scan.results)){
        finite.thresh <- fdr_summaries[ph,which(is.finite(fdr_summaries[ph,]))]
        peak.tables <- lapply()
    }
}

allele_coef <- function(map.geno, map.pheno, map.k, addcovar = NULL,
    intcovar = NULL, scan.results, lod.threshold, exclude.zeros, 
    chr.lod.file, chr.coef.file, round.digs = 2){

    peak.table <- lapply(1:length(scan.results), 
        function(x) find_peaks(scan.results[[x]], map = map, 
        threshold = lod.threshold[x]))

    #round genotypes to reduce risk of low-variance alleles
    #leading to large QTL LOD scores
    rounded.geno <- map.geno
    for(chr in 1:length(rounded.geno)){
        rounded.geno[[chr]] <- round(map.geno[[chr]], round.digs)
    }

    if(!file.exists(chr.lod.file) || overwrite_previous){
        chr.lod <- chr.coef <- vector(mode = "list", length = nrow(Reduce("rbind", peak.table)))
        idx = 1
        for(ph in 1:length(peak.table)){
            if(nrow(peak.table[[ph]]) != 0){
                for(p in 1:nrow(peak.table[[ph]])){
                    chr <- peak.table[[ph]][p,"chr"]
                    peak.loc <- peak.table[[ph]][p,"pos"]
                    trait = names(scan.results)[ph]
                    chr.locale <- which(names(map.geno) == chr)

                    if(exclude.zeros){
                        nonzero.idx <- which(map.pheno[,ph] != 0)
                    }else{
                        nonzero.idx <- 1:nrow(map.pheno)
                    }

                    chr.scan <- scan1(rounded.geno[,chr], 
                        map.pheno[nonzero.idx,ph,drop=FALSE], 
                        kinship = map.k[[chr.locale]], addcovar = addcovar,
                        intcovar = intcovar)
                    chr.lod[[idx]] <- chr.scan
                    #plot(chr.scan, map = map);abline(v = peak.loc)

                    effects.scan <- scan1blup(rounded.geno[,chr], 
                        map.pheno[nonzero.idx,ph,drop=FALSE], 
                        kinship = map.k[[chr.locale]], addcovar = addcovar)
                    chr.coef[[idx]] <- effects.scan
                    #plot_coefCC(effects.scan, map = map);abline(v = peak.loc)
                    idx = idx + 1 #increment regardless of whether there is a peak here
                }
                
            }
        }
        saveRDS(chr.lod, chr.lod.file)
        saveRDS(chr.coef, chr.coef.file)
    }else{
        chr.lod <- readRDS(chr.lod.file)
        chr.coef <- readRDS(chr.coef.file)
    }
    result <- list("peak.table" = peak.table, "LOD" = chr.lod, "Coef" = chr.coef)
    return(result)
}

plot_coef_scans <- function(coef.results, map.geno, map.pheno, fdr_results = NULL,
    tab.level = "###", sig_lines = c("Genome-wide", "FDR")){
    
    if(!is.null(fdr_results)){
        sig_lines = sig_lines[1]
        line.idx <- grep(sig_lines, colnames(fdr_results))
    }

    peak.table <- coef.results$peak.table
    chr.lod <- coef.results$LOD
    chr.coef <- coef.results$Coef

    idx <- 1
    for(ph in 1:length(peak.table)){
        if(nrow(peak.table[[ph]]) != 0){
            for(p in 1:nrow(peak.table[[ph]])){
                par(mfrow = c(2,1))
                chr <- peak.table[[ph]][p,"chr"]
                trait <- colnames(map.pheno)[ph]
                peak.loc <- peak.table[[ph]][p,"pos"]
                chr.locale <- which(names(map.geno) == chr)
                
                cat(tab.level, trait, "Chr", chr, "\n")

                if(!is.null(fdr_results)){
                    ymax <- max(c(chr.lod[[idx]], fdr_results[ph,line.idx]))
                }else{
                    ymax <- max(chr.lod[[idx]])
                }
                plot(chr.lod[[idx]], map = map, main = paste(trait, "Chr", chr), 
                    ylim = c(0, ymax*1.05))
                if(!is.null(fdr_results)){
                    abline(h = fdr_results[ph,line.idx], lty = ph)
                
                    plot.dim <- par("usr")
                    par(xpd = TRUE)
                    text(x = plot.dim[2], y = fdr_results[ph,line.idx], 
                        labels = gsub(paste0(sig_lines, " "), "", colnames(fdr_results)[line.idx]), 
                        cex = 0.5, adj = 0)
                }
                par(xpd = FALSE)


                plot_coefCC(chr.coef[[idx]], map = map, main = paste(trait, "Chr", chr),
                    col = strain.colors)
                abline(v = peak.loc, lwd = 2)

                idx = idx + 1
                cat("\n\n")
            }
        }
    }
}

```

```{r map_cds, warning = FALSE}

cds.scans <- map_pheno(expanded.genoprobs, mapping.pheno, exp.K, 
    addcovar = covar, exclude.zeros = exclude.zeros,results.file = file.path(results.dir, "scan1_results_cds.RDS"))



if(run_permutations){
    #perm_strata <- names(expanded.idx)
    #names(perm_strata) <- rownames(mapping.pheno)
    #cds.perm <- scan1perm(expanded.genoprobs, mapping.pheno, kinship = exp.K, 
    #    addcovar = covar, perm_strata = perm_strata, n_perm = nperm, cores = detectCores()-1)

    #cds.perm <- scan1perm(expanded.genoprobs, mapping.pheno, kinship = exp.K, 
    #    addcovar = covar, n_perm = nperm, cores = detectCores()-1)

    cds.perm <- perm_scans(expanded.genoprobs, mapping.pheno, exp.K,
        cds.scans, addcovar = covar, groups = names(expanded.idx), 
        exclude.zeros = exclude.zeros, nperm = nperm, 
        results.file = file.path(results.dir, "scan1perm_results_cds_perm_scans.RDS"))
}



```




### CDS by individual LOD traces {.tabset .tabset-fade .tabset-pills}

```{r plot_cds_scans, fig.width = 10, fig.height = 4, warning = FALSE, results = "asis"}
#plot the scans, and FDR results
if(run_permutations){
    #pdf("~/Desktop/test.pdf")
    fdr_summaries_cds <- plot_fdr_fun(scan.results = cds.scans, perm.results = cds.perm, 
        tab.level = "####", maxy = NULL)
    #sig_summaries_cds <- plot_pval_fun(scan.results = cds.scans, perm.results = cds.perm, tab.level = "####")
    #dev.off()
}else{
    for(i in 1:length(cds.scans)){
        cat("####", names(cds.scans)[i], "\n")
        plot(cds.scans[[i]], map = map)
        cat("\n\n")
    }
}
```


### CDS by individual allele effects {.tabset .tabset-fade .tabset-pills}

The following plot shows the allele effects for chromosomes 2, 14, 15, 17, and 19 for Global CDS as well as Average Weight Loss. There are two plots per QTL per chromosome. One 
of the full chromosome, and another that captures as much 
of the QTL as possible without showing the weirdness.

The chromosomes that were selected for analysis are based on peaks identified in either Global CDS and/or Average Weight Loss, specially from the Log Transformation with Batch Correlation processing scheme for Average Weight Loss. Thus, this is repeated for Average Weight Loss under that processing scheme. 

Chromosomal Peaks:

- Chr 2, 15, and 17  = Average Weight Loss Peak from Batch, Log Transformation processing

- Chr 19 = Potential shared peak for both

- Chr 14 = Global CDS Peak Originally found using Batch, Rank Z



```{r global_detail, fig.height = 8, fig.width = 8}

#round genotypes to squash out the wild associations we 
#get with teeny tiny probabilities
rounded.geno <- expanded.genoprobs
for(i in 1:length(rounded.geno)){
    rounded.geno[[i]] <- round(rounded.geno[[i]], 2)
}

# chr = 14
# global.lod <- scan1(rounded.geno[,chr], mapping.pheno[,"Global"], exp.K[[chr]],
#     addcovar = covar)
# global.coef <- scan1coef(rounded.geno[,chr], mapping.pheno[,"Global"], exp.K[[chr]],
#     addcovar = covar)
# 
# plot_coefCC(global.coef, map = map, scan1_output = global.lod,
#     ylim = c(-2, 2), col = strain.colors, main = "Global CDS full Chr 14",
#     top_panel_prop = 0.5)
# 
# 
# plot_coefCC(global.coef, map = map, scan1_output = global.lod,
#     ylim = c(-2, 2), col = strain.colors, xlim = c(95, 125),
#     main = "Global CDS Chr 14 QTL Window", top_panel_prop = 0.5)

chr_list = c(2,14,15,17,19)
interval_list = list(c(5, 170), c(95,125), c(55, 74), c(3, 72), c(44, 57))

for(i in 1:length(chr_list)){


# For Global CDS

global.lod <- scan1(rounded.geno[,chr_list[i]], mapping.pheno[,"Global"], exp.K[[chr_list[i]]],
    addcovar = covar)
global.coef <- scan1coef(rounded.geno[,chr_list[i]], mapping.pheno[,"Global"], exp.K[[chr_list[i]]],
    addcovar = covar)

plot_coefCC(global.coef, map = map, scan1_output = global.lod,
    ylim = c(-2, 2), col = strain.colors, main = paste0("Global CDS full Chr ", chr_list[i]),
    top_panel_prop = 0.5)


plot_coefCC(global.coef, map = map, scan1_output = global.lod,
    ylim = c(-2, 2), col = strain.colors, xlim = interval_list[[i]],
    main = paste0("Global CDS Chr ", chr_list[i],  " QTL Window"), top_panel_prop = 0.5)

# for Average Weight Loss
av_weight_Loss.lod <- scan1(rounded.geno[,chr_list[i]], mapping.pheno[,"Avg.Weight.Loss"], exp.K[[chr_list[i]]],
    addcovar = covar)
av_weight_Loss.coef <- scan1coef(rounded.geno[,chr_list[i]], mapping.pheno[,"Avg.Weight.Loss"], exp.K[[chr_list[i]]],
    addcovar = covar)

plot_coefCC(av_weight_Loss.coef, map = map, scan1_output = av_weight_Loss.lod,
    ylim = c(-2, 2), col = strain.colors, main = paste0("Avg Weight Loss full Chr ", chr_list[i]),
    top_panel_prop = 0.5)


plot_coefCC(av_weight_Loss.coef, map = map, scan1_output = av_weight_Loss.lod,
    ylim = c(-2, 2), col = strain.colors, xlim = interval_list[[i]],
    main = paste0("Avg Weight Loss Chr ", chr_list[i],  " QTL Window"), top_panel_prop = 0.5)

}


```



```{r scan_coef_cds, fig.width = 7, fig.height = 7, results = "asis", eval = FALSE}
cds.coef <- allele_coef(map.geno = expanded.genoprobs, 
    map.pheno = mapping.pheno, map.k = exp.K, addcovar = mapping.covar,
    scan.results = cds.scans,
    lod.threshold = allele.thresh, exclude.zeros = exclude.zeros, 
    chr.lod.file = file.path(results.dir, "Chr.LOD.cds.RDS"), 
    chr.coef.file = file.path(results.dir, "Chr.Coef.cds.RDS"))

#pdf("~/Desktop/test.pdf", width = 7, height = 7)
if(run_permutations){
    plot_coef_scans(coef.results = cds.coef, 
        map.geno = expanded.genoprobs, map.pheno = mapping.pheno, 
        fdr_results = fdr_summaries_cds, tab.level = "####",
        sig_lines = "Genome-wide")
}else{
    plot_coef_scans(coef.results = cds.coef, 
        map.geno = expanded.genoprobs, map.pheno = mapping.pheno, 
        fdr_results = NULL, tab.level = "####",
        sig_lines = "Genome-wide")
}

#dev.off()
```


### LOD intervals {.tabset .tabset-fade .tabset-pills} 

The following tables show LOD peak intervals for the top 3 peaks from 
each trait.

#### CDS

```{r lod_peaks_cds, results = "asis"}
top.n <- 5
for(ph in 1:length(cds.scans)){
    cat("#####", names(cds.scans)[ph], "\n")
    peak.table <- find_peaks(cds.scans[[ph]], map = map, prob = 0.95)
    lod.order <- order(peak.table[,"lod"], decreasing = TRUE)
    top.peaks <- peak.table[lod.order[1:top.n],3:ncol(peak.table)]
    rownames(top.peaks) <- NULL
    print(kable(top.peaks))
    cat("Thresholds:\n")
    print(kable(fdr_summaries_cds[ph,,drop=FALSE]))
    cat("\n\n")
}
```



## Log Transformation

All of the steps taken above to yield LOD traces included doing a rank Z tranformation and batch as a covariate. However, these data processing techniques were chosen specifically for CDS before adding the weight Loss data and may impose unknown affects that Loss the story of our other data. Now we will repeat an abbreviated version of the steps run above while changing the following parameters:

* Correction/batch: none, batch, B6

* Transformation: none, rank z, log


### Batch, Log Transformation Set Up

```{r parameters_and_start_up_batch_log_trans}

# preparing to make new files for the new data processing method
file_added_name = 'batch_log_trans' 

# Using the flexibility from the previous code, these parameters will be updated
final_param = TRUE

if(final_param){
    b6_correction <- FALSE
    batch_as_covar <- TRUE
    rank_norm <- FALSE
    log_trans <- TRUE
}
#===============================================================

if(is.na(b6_correction) || length(b6_correction) == 0){
b6_correction = FALSE; batch_as_covar = FALSE; rank_norm = FALSE
}

if(b6_correction && batch_as_covar){stop("Set either b6_correction or batch_as_covar to TRUE, not both.")}

alpha <- c(0.05, 0.2)

overwrite_previous <- FALSE # make this true whenever you add new data that you want fully analyzed; this WILL take longer and may not work on a low powered machine
add_founder_genotypes <- FALSE

#strains identified by remove_strains are excluded from all
#manipulations including normalization and batch corrections.
#justficiations are as follows:
#CC027 - all but 2 were homozygous for the wrong genotype at the H2 locus
#NOD and 129 - are not used for mapping, and were not included in all
#batches, so we cannot use them for normalization
#remove_strains <- NULL
remove_strains <- c("CC027", "NOD", "129")
#remove_strains <- c("CC051", "CC059") #closely related
#remove_strains <- c("CC051") #pick one of them to keep

cc.color.scheme = "original"
#cc.color.scheme = "new"

#for calculations
exclude.zeros = FALSE #if true, we exclude animals with a CDS of 0
disease.thresh = 0 #maximum disease score considered disease-free

nperm = 1000

if(nperm > 0){
    run_permutations <- TRUE
}else{
    run_permutations <- FALSE
}

if(add_founder_genotypes){
    founders.included = "yes"
}else{
    founders.included = "no"
}

if(length(remove_strains) > 0){
    strains.removed <- paste(remove_strains, collapse = ", ")
}else{
    strains.removed = "none"
}

if(!exclude.zeros){
    zero.text <- "Zeros are included in CDS"
    zero.treatment <- "included with"
}else{
    zero.text <- paste("Animals with scores of", disease.thresh, "and below are excluded from CDS.")
    zero.treatment <- "excluded from"
}

if(b6_correction){
    correction.text <- "Batch corrections were made using B6 CDS."
    correction.label <- "B6-batch"
}
if(batch_as_covar){
    correction.text <- "Batch corrections were done using batch as a covariate."
    correction.label <- "covar-batch"
}
if(!b6_correction && !batch_as_covar){
    correction.text <- "No batch corrections were done."
    correction.label <- "no-batch"
}

if(rank_norm){
    norm.text <- "Rank Z normalization was performed."
    norm.label <- "rank-norm"
}else if (log_trans){
    norm.text <- "Log Transformation was performed"
    norm.label <- 'log-trans'
}else{
    norm.text <- "Rank Z normalization was not performed."
    norm.label <- "no-norm"
}

file.label <- paste(correction.label, norm.label, genotypes, sep = "_")
```

The parameters for this run are as follows:

* Genotypes used: `r genotypes`

* Founders included: `r founders.included`

* CC Strains removed: `r strains.removed`

* Treatment of zeros: `r zero.text`

* Batch correction: `r correction.text`

* Normalization: `r norm.text`

```{r reload_data_batch_log_trans}
if(cc.color.scheme == "original"){
    strain.colors <- CCorigcolors
}else{
    strain.colors <- CCcolors
}

cds <- read.csv(here("Data", paste0("CDS_", cds.file.date, ".csv")), strip.white = TRUE)


#take out any slashes in strain names
cds[,"Strain"]  <- gsub("/", "", cds[,"Strain"], fixed = TRUE)

```

```{r remove_strains_batch_log_trans}
#remove any empty rows that were inserted to visually separate
#the founders from the rest of the strains
cds.data.idx <- which(!is.na(cds[,"ID"]))
cds <- cds[cds.data.idx,]


#remove specified strains from cds
if(length(remove_strains) != 0){
    remove.cds <- unlist(sapply(remove_strains, function(x) which(cds[,"Strain"] == x)))
    cds <- cds[-remove.cds,]
}


```


```{r batches_batch_log_trans}
## CDS
num.cds <- as.matrix(sapply(cds[,c( "Global.CDS", "Avg.Weight.Change","Avg.Weight.Loss", "Weight.D0")], as.numeric)) #had to change this with Chat to make the values numeric
rownames(num.cds) <- cds[,"ID"]
#set all scores equal to or lower than the disease threshold
#to zero
if(exclude.zeros){
    num.cds[which(num.cds <= disease.thresh)] <- 0
}

factor.cds <- as.matrix(cds[,c("Strain", "Sex")])
u_batch <- unique(cds[,"EXP"])

#remove animals in batch 170 - these were validation, not part of the original experiment
to.remove = "X170"
keep.batch <- which(u_batch != "X170")
u_batch <- u_batch[keep.batch]

keep.idx <- which(cds[,"EXP"] != "X170")
cds <- cds[keep.idx,]
num.cds <- num.cds[keep.idx,]
factor.cds <- factor.cds[keep.idx,]

batch.mat <- matrix(0, nrow = nrow(cds), ncol = length(u_batch))
colnames(batch.mat) <- u_batch
for(b in 1:ncol(batch.mat)){
    batch.mat[which(cds[,"EXP"] == u_batch[b]),b] <- 1
}
cds.covar <- as.matrix(cbind(batch.mat, cds[,"Age.at.D0",drop=FALSE]))

pheno.strains <- cds[,"Strain"]
eae.strains <- unique(pheno.strains)


```


```{r adj_cds_batch_log_trans}
#choose which phenotypes to batch
#correct using their indices.
cds.idx <- c(1,2,3,4)
adj.cds <- num.cds

if(batch_as_covar){
    adj.cds <- adjust(num.cds, cds.covar[,1:ncol(cds.covar)-1])
}

ymin = NULL; ymax = NULL #all on independend axes
for(cd in 1:ncol(norm.cds)){
    cat("####", colnames(norm.cds)[cd], "\n")
    plot_qtl_by_strain(scores = norm.cds[,cd], 
        strain.labels = factor.cds[,"Strain"], exclude.zeros = exclude.zeros,
        ymin = ymin, ymax = ymax, plot.label = colnames(num.cds)[cd])
    cat("\n\n")
}
```



```{r adjust_age_cds_batch_log_trans, eval = FALSE}
#adjust for age
for(i in 1:ncol(adj.cds)){
    if(exclude.zeros){
        nonzero.idx <- which(num.cds[,i] != 0)
    }else{
        nonzero.idx <- 1:nrow(num.cds)
    }
    adj.cds[nonzero.idx,i] <- adjust(adj.cds[nonzero.idx,i,drop=FALSE],
        cds.covar[nonzero.idx,"Age.at.D0",drop=FALSE])
}

#write out the results of whatever adjustments we have done
write.table(adj.cds, file.path(results.dir, paste0("Adj_CDS", file_added_name, ".csv")), sep = ",", quote = FALSE)

ymin = NULL; ymax = NULL #all on independend axes
for(cd in 1:ncol(norm.cds)){
    cat("####", colnames(norm.cds)[cd], "\n")
    plot_qtl_by_strain(scores = norm.cds[,cd], 
        strain.labels = factor.cds[,"Strain"], exclude.zeros = exclude.zeros,
        ymin = ymin, ymax = ymax, plot.label = colnames(num.cds)[cd])
    cat("\n\n")
}
```

```{r no_norm_cds_batch_log_trans}
norm.cds <- adj.cds
colnames(norm.cds) <- gsub(".CDS", "", colnames(norm.cds)) #take off CDS in labels, to save room in plots

if(rank_norm){
    for(i in 1:ncol(norm.cds)){
        if(exclude.zeros){
            nonzero.idx <- which(num.cds[,i] != 0)
        }else{
            nonzero.idx <- 1:nrow(num.cds)
        }
        norm.cds[nonzero.idx,i] <- rankZ(adj.cds[nonzero.idx,i])
    }
}

if(log_trans){
    for(i in 1:ncol(norm.cds)){
        if(exclude.zeros){
            nonzero.idx <- which(num.cds[,i] != 0)
        }else{
            nonzero.idx <- 1:nrow(num.cds)
        }
      if (min(adj.cds[,i])<0){
        adj.cds <- adj.cds - min(adj.cds[,i])
      }  
        norm.cds[nonzero.idx,i] <- log(adj.cds[nonzero.idx,i]+1)
    }
}
colnames(norm.cds) <- gsub(".CDS", "", colnames(norm.cds)) #take off CDS, to save room in plots

#write out normzliaed values
write.table(norm.cds, file.path(results.dir, paste0("Norm_CDS",file_added_name,".csv")), sep = ",", quote = FALSE)

```



```{r methods_batch_log_trans}
if(add_founder_genotypes){
    methods.line <- "Both CC and founder strains were used for mapping."
}else{
    methods.line <- "We used only CC strains for mapping."
}

if(length(remove_strains) != 0){
    remove.line <- paste("We excluded the following strains:", paste(remove_strains, collapse = ", "))
}else{
    remove.line <- "No CC strains were excluded."
}


```

`r methods.line`
`r remove.line`

### Normalized Scores {.tabset .tabset-fade .tabset-pills}

This is post processing
```{r norm_dist_CDS_log_trans, fig.width = 10, fig.height = 5, results = "asis"}
#ymin = min(norm.cds);ymax = max(norm.cds) #all on same axes
ymin = NULL; ymax = NULL #all on independend axes
for(cd in 1:ncol(norm.cds)){
    cat("####", colnames(norm.cds)[cd], "\n")
    plot_qtl_by_strain(scores = norm.cds[,cd], 
        strain.labels = factor.cds[,"Strain"], exclude.zeros = exclude.zeros,
        ymin = ymin, ymax = ymax, plot.label = colnames(num.cds)[cd])
    cat("\n\n")
}

```

### Overall Sex Effect

```{r sex_diff_cds_batch_log_trans}
#quartz(width = 8, height = 8)
par(mfrow = c(2,2))
for(i in 1:ncol(norm.cds)){
    if(exclude.zeros){
        nonzero.idx <- which(norm.cds[,i] != 0)
    }else{
        nonzero.idx <- 1:nrow(norm.cds)
    }
    model <- lm(norm.cds[nonzero.idx,i]~as.factor(factor.cds[nonzero.idx,"Sex"]))
    f <- summary(model)$fstatistic
    p <- pf(f[1],f[2],f[3],lower.tail=F)
    boxplot(norm.cds[nonzero.idx,i]~as.factor(factor.cds[nonzero.idx,"Sex"]), 
        ylab = colnames(norm.cds)[i],
        main = paste(colnames(norm.cds)[i], "\np =", signif(p, 2)), xlab = "")
}
```



### Trait Correlations

The following plot shows that global CDS
is dominated by classical disease.
It also shows that CDS in classical 
disease and AR are unrelated to each other. 
Lastly, it shows the correlations between the
average weight Loss and the CDS data which are negative. This makes sense because weight loss is associated with higher disease severity.

```{r trait_cor_cds_log_trans, fig.width = 8, fig.height = 12}
trait.pairs <- pair.matrix(1:ncol(norm.cds))
par(mfrow = c(3,2))
for(i in 1:nrow(trait.pairs)){
    trait1 <- trait.pairs[i,1]
    trait2 <- trait.pairs[i,2]
    if(exclude.zeros){
        nonzero.idx <- intersect(which(norm.cds[,trait1] != 0), which(norm.cds[,trait2] != 0))
    }else{
        nonzero.idx <- 1:nrow(norm.cds)
    }
    plot.with.model(norm.cds[nonzero.idx,trait1], norm.cds[nonzero.idx,trait2],
        xlab = colnames(norm.cds)[trait1], ylab = colnames(norm.cds)[trait2],
        main = paste(colnames(norm.cds)[trait1], "vs.", colnames(norm.cds)[trait2]))
}
```

```{r expand_matrices_cds_batch_log_trans}

#remove any strains not used in mapping
mapping.strain.idx <- unlist(sapply(mapping.strains, function(x) grep(x, pheno.strains)))
pheno.strains_cds <- pheno.strains[mapping.strain.idx]

#make a unique ID for each mouse that contains its strain and sex
u_id <- as.vector(apply(cds[mapping.strain.idx,c(1:2,5)], 1, function(x) paste(x, collapse = "_")))

K.idx <- sapply(mapping.strains, function(x) grep(x, rownames(mapping.K[[1]])))
K.num <- sapply(mapping.strains, function(x) length(which(pheno.strains_cds == x)))
expanded.idx <- unlist(lapply(1:length(K.idx), function(x) rep(K.idx[x], K.num[x])))
groups <- names(expanded.idx)

exp.K <- lapply(1:length(mapping.K), function(x) mapping.K[[x]][expanded.idx,expanded.idx])
names(exp.K) <- names(mapping.K)

#create unique row and column names
for(i in 1:length(exp.K)){
    rownames(exp.K[[i]]) <- colnames(exp.K[[i]]) <- 1:length(u_id)
}

expanded.genoprobs <- mapping.genoprobs
for(i in 1:length(expanded.genoprobs)){
    expanded.genoprobs[[i]] <- mapping.genoprobs[[i]][expanded.idx,,]
    rownames(expanded.genoprobs[[i]]) <- 1:length(u_id)
}

#align the phenotypes with the strains
covar <- matrix(as.numeric(as.factor(cds[mapping.strain.idx,"Sex"])), ncol = 1)
rownames(covar) <- 1:length(u_id)

mapping.pheno <- norm.cds[mapping.strain.idx,]
rownames(mapping.pheno) <- 1:length(u_id)

mapping.factors <- factor.cds[mapping.strain.idx,]
mapping.covar <- matrix(as.numeric(as.factor(mapping.factors[,"Sex"])), ncol = 1)
rownames(mapping.covar) <- 1:length(u_id)

```

### Heritability

```{r herit_cds_batch_log_trans, fig.width = 7, fig.height = 5}
overall.k.file <- here("Data",paste0("kinship_overall_cds",file_added_name,".RDS"))
if(!file.exists(overall.k.file) || overwrite_previous){
    overall.K <- calc_kinship(expanded.genoprobs, "overall")
    saveRDS(overall.K, overall.k.file)
}else{
    overall.K <- readRDS(overall.k.file)
}

ind.herit <- rep(NA, ncol(num.cds))
names(ind.herit) <- colnames(num.cds)
for(i in 1:ncol(num.cds)){
    nonzero.idx <- which(mapping.pheno[,i] != 0)
    ind.herit[i] <- est_herit(mapping.pheno[nonzero.idx,i,drop=FALSE], overall.K)
}

barplot_with_num(round(ind.herit, 2), ylab = "Heritability", xlab = "", 
    ylim = c(0, 0.4), text.shift = 0.09, main = "Trait Heritability Estimates")
```




```{r map_cds_batch_log_trans, warning = FALSE}

cds.scans <- map_pheno(expanded.genoprobs, mapping.pheno, exp.K,
    addcovar = covar, exclude.zeros = exclude.zeros,results.file = file.path(results.dir, paste0("scan1_results_cds",file_added_name,".RDS")))



if(run_permutations){
    #perm_strata <- names(expanded.idx)
    #names(perm_strata) <- rownames(mapping.pheno)
    #cds.perm <- scan1perm(expanded.genoprobs, mapping.pheno, kinship = exp.K,
    #    addcovar = covar, perm_strata = perm_strata, n_perm = nperm, cores = detectCores()-1)

    #cds.perm <- scan1perm(expanded.genoprobs, mapping.pheno, kinship = exp.K,
    #    addcovar = covar, n_perm = nperm, cores = detectCores()-1)

    cds.perm <- perm_scans(expanded.genoprobs, mapping.pheno, exp.K,
        cds.scans, addcovar = covar, groups = names(expanded.idx),
        exclude.zeros = exclude.zeros, nperm = nperm,
        results.file = file.path(results.dir, paste0("scan1perm_results_cds_perm_scans",file_added_name,".RDS")))
}
```

### CDS by individual LOD traces for Batch, Log Transformation {.tabset .tabset-fade .tabset-pills}

```{r plot_cds_scans_batch_log_trans, fig.width = 10, fig.height = 4, warning = FALSE, results = "asis"}
#plot the scans, and FDR results
if(run_permutations){
    #pdf("~/Desktop/test.pdf")
    fdr_summaries_cds <- plot_fdr_fun(scan.results = cds.scans, perm.results = cds.perm,
        tab.level = "#####", maxy = NULL)
    #sig_summaries_cds <- plot_pval_fun(scan.results = cds.scans, perm.results = cds.perm, tab.level = "####")
    #dev.off()
}else{
    for(i in 1:length(cds.scans)){
        cat("#####", names(cds.scans)[i], "\n")
        plot(cds.scans[[i]], map = map)
        cat("\n\n")
    }
}
```



### LOD intervals for Batch, Log Transformation {.tabset .tabset-fade .tabset-pills}

```{r lod_peaks_cds_batch_log_trans, results = "asis"}
top.n <- 5
for(ph in 1:length(cds.scans)){
    cat("#####", names(cds.scans)[ph], "\n")
    peak.table <- find_peaks(cds.scans[[ph]], map = map, prob = 0.95)
    lod.order <- order(peak.table[,"lod"], decreasing = TRUE)
    top.peaks <- peak.table[lod.order[1:top.n],3:ncol(peak.table)]
    rownames(top.peaks) <- NULL
    print(kable(top.peaks))
    cat("Thresholds:\n")
    print(kable(fdr_summaries_cds[ph,,drop=FALSE]))
    cat("\n\n")
}
```



#### Average Weight Loss by individual allele effects {.tabset .tabset-fade .tabset-pills}

The following plot shows the allele effects for the Chr 2
QTL for Average Weight Loss. The methods for plotting this
chromosome and phenotype are the same done for Global CDS previously, including
the two separate plots with x-limit perimeters modified 
as appropriate.

```{r av_weight_Loss_detail, fig.height = 8, fig.width = 8}

#round genotypes to squash out the wild associations we 
#get with teeny tiny probabilities
rounded.geno <- expanded.genoprobs
for(i in 1:length(rounded.geno)){
    rounded.geno[[i]] <- round(rounded.geno[[i]], 2)
}

# chr = 2
# av_weight_Loss.lod <- scan1(rounded.geno[,chr], mapping.pheno[,"Avg.Weight.Loss"], exp.K[[chr]],
#     addcovar = covar)
# av_weight_Loss.coef <- scan1coef(rounded.geno[,chr], mapping.pheno[,"Avg.Weight.Loss"], exp.K[[chr]],
#     addcovar = covar)
# 
# plot_coefCC(av_weight_Loss.coef, map = map, scan1_output = av_weight_Loss.lod,
#     ylim = c(-2, 2), col = strain.colors, main = "Avg Weight Loss full Chr 2",
#     top_panel_prop = 0.5)
# 
# 
# plot_coefCC(av_weight_Loss.coef, map = map, scan1_output = av_weight_Loss.lod,
#     ylim = c(-2, 2), col = strain.colors, xlim = c(145, 180),
#     main = "Avg Weight Loss Chr 2 QTL Window", top_panel_prop = 0.5)

chr_list = c(2,14,15,17,19)
interval_list = list(c(125, 175), c(95,125), c(55, 74), c(3, 72), c(44, 57))

for(i in 1:length(chr_list)){

# for Average Weight Loss
av_weight_Loss.lod <- scan1(rounded.geno[,chr_list[i]], mapping.pheno[,"Avg.Weight.Loss"], exp.K[[chr_list[i]]],
    addcovar = covar)
av_weight_Loss.coef <- scan1coef(rounded.geno[,chr_list[i]], mapping.pheno[,"Avg.Weight.Loss"], exp.K[[chr_list[i]]],
    addcovar = covar)

plot_coefCC(av_weight_Loss.coef, map = map, scan1_output = av_weight_Loss.lod,
    ylim = c(-2, 2), col = strain.colors, main = paste0("Avg Weight Loss full Chr ", chr_list[i]),
    top_panel_prop = 0.5)


plot_coefCC(av_weight_Loss.coef, map = map, scan1_output = av_weight_Loss.lod,
    ylim = c(-2, 2), col = strain.colors, xlim = interval_list[[i]],
    main = paste0("Avg Weight Loss Chr ", chr_list[i],  " QTL Window"), top_panel_prop = 0.5)
}

```


## Log Transformation with Shifted Data

Instead of transforming the original weight change and loss data which both have negative values, this will shift the data to start at zero.


### Batch, Log Transformation Set Up

```{r parameters_and_start_up_batch_log_trans_shift}

# preparing to make new files for the new data processing method
file_added_name = 'batch_log_trans_shift' 

# Using the flexibility from the previous code, these parameters will be updated
final_param = TRUE

if(final_param){
    b6_correction <- FALSE
    batch_as_covar <- TRUE
    rank_norm <- FALSE
    log_trans <- TRUE
}
#===============================================================

if(is.na(b6_correction) || length(b6_correction) == 0){
b6_correction = FALSE; batch_as_covar = FALSE; rank_norm = FALSE
}

if(b6_correction && batch_as_covar){stop("Set either b6_correction or batch_as_covar to TRUE, not both.")}

alpha <- c(0.05, 0.2)

overwrite_previous <- FALSE # make this true whenever you add new data that you want fully analyzed; this WILL take longer and may not work on a low powered machine
add_founder_genotypes <- FALSE

#strains identified by remove_strains are excluded from all
#manipulations including normalization and batch corrections.
#justficiations are as follows:
#CC027 - all but 2 were homozygous for the wrong genotype at the H2 locus
#NOD and 129 - are not used for mapping, and were not included in all
#batches, so we cannot use them for normalization
#remove_strains <- NULL
remove_strains <- c("CC027", "NOD", "129")
#remove_strains <- c("CC051", "CC059") #closely related
#remove_strains <- c("CC051") #pick one of them to keep

cc.color.scheme = "original"
#cc.color.scheme = "new"

#for calculations
exclude.zeros = FALSE #if true, we exclude animals with a CDS of 0
disease.thresh = 0 #maximum disease score considered disease-free

nperm = 1000

if(nperm > 0){
    run_permutations <- TRUE
}else{
    run_permutations <- FALSE
}

if(add_founder_genotypes){
    founders.included = "yes"
}else{
    founders.included = "no"
}

if(length(remove_strains) > 0){
    strains.removed <- paste(remove_strains, collapse = ", ")
}else{
    strains.removed = "none"
}

if(!exclude.zeros){
    zero.text <- "Zeros are included in CDS"
    zero.treatment <- "included with"
}else{
    zero.text <- paste("Animals with scores of", disease.thresh, "and below are excluded from CDS.")
    zero.treatment <- "excluded from"
}

if(b6_correction){
    correction.text <- "Batch corrections were made using B6 CDS."
    correction.label <- "B6-batch"
}
if(batch_as_covar){
    correction.text <- "Batch corrections were done using batch as a covariate."
    correction.label <- "covar-batch"
}
if(!b6_correction && !batch_as_covar){
    correction.text <- "No batch corrections were done."
    correction.label <- "no-batch"
}

if(rank_norm){
    norm.text <- "Rank Z normalization was performed."
    norm.label <- "rank-norm"
}else if (log_trans){
    norm.text <- "Log Transformation was performed"
    norm.label <- 'log-trans'
}else{
    norm.text <- "Rank Z normalization was not performed."
    norm.label <- "no-norm"
}

file.label <- paste(correction.label, norm.label, genotypes, sep = "_")
```

The parameters for this run are as follows:

* Genotypes used: `r genotypes`

* Founders included: `r founders.included`

* CC Strains removed: `r strains.removed`

* Treatment of zeros: `r zero.text`

* Batch correction: `r correction.text`

* Normalization: `r norm.text`

```{r reload_data_batch_log_trans_shift}
if(cc.color.scheme == "original"){
    strain.colors <- CCorigcolors
}else{
    strain.colors <- CCcolors
}

cds <- read.csv(here("Data", paste0("CDS_", cds.file.date, ".csv")), strip.white = TRUE)


#take out any slashes in strain names
cds[,"Strain"]  <- gsub("/", "", cds[,"Strain"], fixed = TRUE)

```

```{r remove_strains_batch_log_trans_shift}
#remove any empty rows that were inserted to visually separate
#the founders from the rest of the strains
cds.data.idx <- which(!is.na(cds[,"ID"]))
cds <- cds[cds.data.idx,]


#remove specified strains from cds
if(length(remove_strains) != 0){
    remove.cds <- unlist(sapply(remove_strains, function(x) which(cds[,"Strain"] == x)))
    cds <- cds[-remove.cds,]
}


```


```{r batches_batch_log_trans_shift}
## CDS
num.cds <- as.matrix(sapply(cds[,c( "Global.CDS", "Avg.Weight.Change","Avg.Weight.Loss", "Weight.D0")], as.numeric)) #had to change this with Chat to make the values numeric
rownames(num.cds) <- cds[,"ID"]

# shift all values so that the lowest score is zero
for(colnum in 1:ncol(num.cds)){
  num.cds[,colnum] = num.cds[,colnum] + abs(min((num.cds[,colnum]), na.rm = TRUE))
}


#set all scores equal to or lower than the disease threshold
#to zero
if(exclude.zeros){
    num.cds[which(num.cds <= disease.thresh)] <- 0
}

factor.cds <- as.matrix(cds[,c("Strain", "Sex")])
u_batch <- unique(cds[,"EXP"])

#remove animals in batch 170 - these were validation, not part of the original experiment
to.remove = "X170"
keep.batch <- which(u_batch != "X170")
u_batch <- u_batch[keep.batch]

keep.idx <- which(cds[,"EXP"] != "X170")
cds <- cds[keep.idx,]
num.cds <- num.cds[keep.idx,]
factor.cds <- factor.cds[keep.idx,]

batch.mat <- matrix(0, nrow = nrow(cds), ncol = length(u_batch))
colnames(batch.mat) <- u_batch
for(b in 1:ncol(batch.mat)){
    batch.mat[which(cds[,"EXP"] == u_batch[b]),b] <- 1
}
cds.covar <- as.matrix(cbind(batch.mat, cds[,"Age.at.D0",drop=FALSE]))

pheno.strains <- cds[,"Strain"]
eae.strains <- unique(pheno.strains)


```


```{r adj_cds_batch_log_trans_shift}
#choose which phenotypes to batch
#correct using their indices.
cds.idx <- c(1,2,3,4)
adj.cds <- num.cds

if(batch_as_covar){
    adj.cds <- adjust(num.cds, cds.covar[,1:ncol(cds.covar)-1])
}
```



```{r adjust_age_cds_batch_log_trans_shift, eval = FALSE}
#adjust for age
for(i in 1:ncol(adj.cds)){
    if(exclude.zeros){
        nonzero.idx <- which(num.cds[,i] != 0)
    }else{
        nonzero.idx <- 1:nrow(num.cds)
    }
    adj.cds[nonzero.idx,i] <- adjust(adj.cds[nonzero.idx,i,drop=FALSE],
        cds.covar[nonzero.idx,"Age.at.D0",drop=FALSE])
}

#write out the results of whatever adjustments we have done
write.table(adj.cds, file.path(results.dir, paste0("Adj_CDS", file_added_name, ".csv")), sep = ",", quote = FALSE)
```

```{r no_norm_cds_batch_log_trans_shift}
norm.cds <- adj.cds
colnames(norm.cds) <- gsub(".CDS", "", colnames(norm.cds)) #take off CDS in labels, to save room in plots

if(rank_norm){
    for(i in 1:ncol(norm.cds)){
        if(exclude.zeros){
            nonzero.idx <- which(num.cds[,i] != 0)
        }else{
            nonzero.idx <- 1:nrow(num.cds)
        }
        norm.cds[nonzero.idx,i] <- rankZ(adj.cds[nonzero.idx,i])
    }
}

if(log_trans){
    for(i in 1:ncol(norm.cds)){
        if(exclude.zeros){
            nonzero.idx <- which(num.cds[,i] != 0)
        }else{
            nonzero.idx <- 1:nrow(num.cds)
        }
      if (min(adj.cds[,i])<0){
        adj.cds <- adj.cds - min(adj.cds[,i])
      }  
        norm.cds[nonzero.idx,i] <- log(adj.cds[nonzero.idx,i]+1)
    }
}
colnames(norm.cds) <- gsub(".CDS", "", colnames(norm.cds)) #take off CDS, to save room in plots

#write out normzliaed values
write.table(norm.cds, file.path(results.dir, paste0("Norm_CDS",file_added_name,".csv")), sep = ",", quote = FALSE)

```



```{r methods_batch_log_trans_shift}
if(add_founder_genotypes){
    methods.line <- "Both CC and founder strains were used for mapping."
}else{
    methods.line <- "We used only CC strains for mapping."
}

if(length(remove_strains) != 0){
    remove.line <- paste("We excluded the following strains:", paste(remove_strains, collapse = ", "))
}else{
    remove.line <- "No CC strains were excluded."
}


```

`r methods.line`
`r remove.line`

### Normalized Scores {.tabset .tabset-fade .tabset-pills}

This is post processing
```{r norm_dist_CDS_log_trans_shift, fig.width = 10, fig.height = 5, results = "asis"}
#ymin = min(norm.cds);ymax = max(norm.cds) #all on same axes
ymin = NULL; ymax = NULL #all on independend axes
for(cd in 1:ncol(norm.cds)){
    cat("####", colnames(norm.cds)[cd], "\n")
    plot_qtl_by_strain(scores = norm.cds[,cd], 
        strain.labels = factor.cds[,"Strain"], exclude.zeros = exclude.zeros,
        ymin = ymin, ymax = ymax, plot.label = colnames(num.cds)[cd])
    cat("\n\n")
}

```

### Overall Sex Effect

```{r sex_diff_cds_batch_log_trans_shift}
#quartz(width = 8, height = 8)
par(mfrow = c(2,2))
for(i in 1:ncol(norm.cds)){
    if(exclude.zeros){
        nonzero.idx <- which(norm.cds[,i] != 0)
    }else{
        nonzero.idx <- 1:nrow(norm.cds)
    }
    model <- lm(norm.cds[nonzero.idx,i]~as.factor(factor.cds[nonzero.idx,"Sex"]))
    f <- summary(model)$fstatistic
    p <- pf(f[1],f[2],f[3],lower.tail=F)
    boxplot(norm.cds[nonzero.idx,i]~as.factor(factor.cds[nonzero.idx,"Sex"]), 
        ylab = colnames(norm.cds)[i],
        main = paste(colnames(norm.cds)[i], "\np =", signif(p, 2)), xlab = "")
}
```



### Trait Correlations

The following plot shows that global CDS
is dominated by classical disease.
It also shows that CDS in classical 
disease and AR are unrelated to each other. 
Lastly, it shows the correlations between the
average weight Loss and the CDS data which are negative. This makes sense because weight loss is associated with higher disease severity.

```{r trait_cor_cds_log_trans_shift, fig.width = 8, fig.height = 12}
trait.pairs <- pair.matrix(1:ncol(norm.cds))
par(mfrow = c(3,2))
for(i in 1:nrow(trait.pairs)){
    trait1 <- trait.pairs[i,1]
    trait2 <- trait.pairs[i,2]
    if(exclude.zeros){
        nonzero.idx <- intersect(which(norm.cds[,trait1] != 0), which(norm.cds[,trait2] != 0))
    }else{
        nonzero.idx <- 1:nrow(norm.cds)
    }
    plot.with.model(norm.cds[nonzero.idx,trait1], norm.cds[nonzero.idx,trait2],
        xlab = colnames(norm.cds)[trait1], ylab = colnames(norm.cds)[trait2],
        main = paste(colnames(norm.cds)[trait1], "vs.", colnames(norm.cds)[trait2]))
}
```

```{r expand_matrices_cds_batch_log_trans_shift}

#remove any strains not used in mapping
mapping.strain.idx <- unlist(sapply(mapping.strains, function(x) grep(x, pheno.strains)))
pheno.strains_cds <- pheno.strains[mapping.strain.idx]

#make a unique ID for each mouse that contains its strain and sex
u_id <- as.vector(apply(cds[mapping.strain.idx,c(1:2,5)], 1, function(x) paste(x, collapse = "_")))

K.idx <- sapply(mapping.strains, function(x) grep(x, rownames(mapping.K[[1]])))
K.num <- sapply(mapping.strains, function(x) length(which(pheno.strains_cds == x)))
expanded.idx <- unlist(lapply(1:length(K.idx), function(x) rep(K.idx[x], K.num[x])))
groups <- names(expanded.idx)

exp.K <- lapply(1:length(mapping.K), function(x) mapping.K[[x]][expanded.idx,expanded.idx])
names(exp.K) <- names(mapping.K)

#create unique row and column names
for(i in 1:length(exp.K)){
    rownames(exp.K[[i]]) <- colnames(exp.K[[i]]) <- 1:length(u_id)
}

expanded.genoprobs <- mapping.genoprobs
for(i in 1:length(expanded.genoprobs)){
    expanded.genoprobs[[i]] <- mapping.genoprobs[[i]][expanded.idx,,]
    rownames(expanded.genoprobs[[i]]) <- 1:length(u_id)
}

#align the phenotypes with the strains
covar <- matrix(as.numeric(as.factor(cds[mapping.strain.idx,"Sex"])), ncol = 1)
rownames(covar) <- 1:length(u_id)

mapping.pheno <- norm.cds[mapping.strain.idx,]
rownames(mapping.pheno) <- 1:length(u_id)

mapping.factors <- factor.cds[mapping.strain.idx,]
mapping.covar <- matrix(as.numeric(as.factor(mapping.factors[,"Sex"])), ncol = 1)
rownames(mapping.covar) <- 1:length(u_id)

```

### Heritability

```{r herit_cds_batch_log_trans_shift, fig.width = 7, fig.height = 5}
overall.k.file <- here("Data",paste0("kinship_overall_cds",file_added_name,".RDS"))
if(!file.exists(overall.k.file) || overwrite_previous){
    overall.K <- calc_kinship(expanded.genoprobs, "overall")
    saveRDS(overall.K, overall.k.file)
}else{
    overall.K <- readRDS(overall.k.file)
}

ind.herit <- rep(NA, ncol(num.cds))
names(ind.herit) <- colnames(num.cds)
for(i in 1:ncol(num.cds)){
    nonzero.idx <- which(mapping.pheno[,i] != 0)
    ind.herit[i] <- est_herit(mapping.pheno[nonzero.idx,i,drop=FALSE], overall.K)
}

barplot_with_num(round(ind.herit, 2), ylab = "Heritability", xlab = "", 
    ylim = c(0, 0.4), text.shift = 0.09, main = "Trait Heritability Estimates")
```




```{r map_cds_batch_log_trans_shift, warning = FALSE}

cds.scans <- map_pheno(expanded.genoprobs, mapping.pheno, exp.K,
    addcovar = covar, exclude.zeros = exclude.zeros,results.file = file.path(results.dir, paste0("scan1_results_cds",file_added_name,".RDS")))



if(run_permutations){
    #perm_strata <- names(expanded.idx)
    #names(perm_strata) <- rownames(mapping.pheno)
    #cds.perm <- scan1perm(expanded.genoprobs, mapping.pheno, kinship = exp.K,
    #    addcovar = covar, perm_strata = perm_strata, n_perm = nperm, cores = detectCores()-1)

    #cds.perm <- scan1perm(expanded.genoprobs, mapping.pheno, kinship = exp.K,
    #    addcovar = covar, n_perm = nperm, cores = detectCores()-1)

    cds.perm <- perm_scans(expanded.genoprobs, mapping.pheno, exp.K,
        cds.scans, addcovar = covar, groups = names(expanded.idx),
        exclude.zeros = exclude.zeros, nperm = nperm,
        results.file = file.path(results.dir, paste0("scan1perm_results_cds_perm_scans",file_added_name,".RDS")))
}
```

### CDS by individual LOD traces for Batch, Log Transformation {.tabset .tabset-fade .tabset-pills}

```{r plot_cds_scans_batch_log_trans_shift, fig.width = 10, fig.height = 4, warning = FALSE, results = "asis"}
#plot the scans, and FDR results
if(run_permutations){
    #pdf("~/Desktop/test.pdf")
    fdr_summaries_cds <- plot_fdr_fun(scan.results = cds.scans, perm.results = cds.perm,
        tab.level = "#####", maxy = NULL)
    #sig_summaries_cds <- plot_pval_fun(scan.results = cds.scans, perm.results = cds.perm, tab.level = "####")
    #dev.off()
}else{
    for(i in 1:length(cds.scans)){
        cat("#####", names(cds.scans)[i], "\n")
        plot(cds.scans[[i]], map = map)
        cat("\n\n")
    }
}
```



### LOD intervals for Batch, Log Transformation {.tabset .tabset-fade .tabset-pills}

```{r lod_peaks_cds_batch_log_trans_shift, results = "asis"}
top.n <- 5
for(ph in 1:length(cds.scans)){
    cat("#####", names(cds.scans)[ph], "\n")
    peak.table <- find_peaks(cds.scans[[ph]], map = map, prob = 0.95)
    lod.order <- order(peak.table[,"lod"], decreasing = TRUE)
    top.peaks <- peak.table[lod.order[1:top.n],3:ncol(peak.table)]
    rownames(top.peaks) <- NULL
    print(kable(top.peaks))
    cat("Thresholds:\n")
    print(kable(fdr_summaries_cds[ph,,drop=FALSE]))
    cat("\n\n")
}
```



#### Average Weight Loss by individual allele effects {.tabset .tabset-fade .tabset-pills}

The following plot shows the allele effects for the Chr 2
QTL for Average Weight Loss. The methods for plotting this
chromosome and phenotype are the same done for Global CDS previously, including
the two separate plots with x-limit perimeters modified 
as appropriate.

```{r av_weight_Loss_detail_shift, fig.height = 8, fig.width = 8}

#round genotypes to squash out the wild associations we 
#get with teeny tiny probabilities
rounded.geno <- expanded.genoprobs
for(i in 1:length(rounded.geno)){
    rounded.geno[[i]] <- round(rounded.geno[[i]], 2)
}

# chr = 2
# av_weight_Loss.lod <- scan1(rounded.geno[,chr], mapping.pheno[,"Avg.Weight.Loss"], exp.K[[chr]],
#     addcovar = covar)
# av_weight_Loss.coef <- scan1coef(rounded.geno[,chr], mapping.pheno[,"Avg.Weight.Loss"], exp.K[[chr]],
#     addcovar = covar)
# 
# plot_coefCC(av_weight_Loss.coef, map = map, scan1_output = av_weight_Loss.lod,
#     ylim = c(-2, 2), col = strain.colors, main = "Avg Weight Loss full Chr 2",
#     top_panel_prop = 0.5)
# 
# 
# plot_coefCC(av_weight_Loss.coef, map = map, scan1_output = av_weight_Loss.lod,
#     ylim = c(-2, 2), col = strain.colors, xlim = c(145, 180),
#     main = "Avg Weight Loss Chr 2 QTL Window", top_panel_prop = 0.5)

chr_list = c(2,14,15,17,19)
interval_list = list(c(125, 175), c(95,125), c(55, 74), c(3, 72), c(44, 57))

for(i in 1:length(chr_list)){

# for Average Weight Loss
av_weight_Loss.lod <- scan1(rounded.geno[,chr_list[i]], mapping.pheno[,"Avg.Weight.Loss"], exp.K[[chr_list[i]]],
    addcovar = covar)
av_weight_Loss.coef <- scan1coef(rounded.geno[,chr_list[i]], mapping.pheno[,"Avg.Weight.Loss"], exp.K[[chr_list[i]]],
    addcovar = covar)

plot_coefCC(av_weight_Loss.coef, map = map, scan1_output = av_weight_Loss.lod,
    ylim = c(-2, 2), col = strain.colors, main = paste0("Avg Weight Loss full Chr ", chr_list[i]),
    top_panel_prop = 0.5)


plot_coefCC(av_weight_Loss.coef, map = map, scan1_output = av_weight_Loss.lod,
    ylim = c(-2, 2), col = strain.colors, xlim = interval_list[[i]],
    main = paste0("Avg Weight Loss Chr ", chr_list[i],  " QTL Window"), top_panel_prop = 0.5)
}

```